Resolvable Ambiguity

### 2019-07-10 paper/paper.tex
Can we? Will we?
Polishing Orc case study (including David's comments)

\synctex=1
\begin{document}

@@

@@

\subsection{Orc} \label{sec:evaluation-orc}

Orc~\cite{kitchinOrc2009} is a functional programming language
which includes a number of special-purpose combinators for
coordinating concurrent workflows. In this section, we consider
the use case of adding these concurrency combinators to an already
defined programming language.

In Orc, every expression can ``publish'' zero or more values. Orc
defines four combinators for orchestrating these published values:
the parallel (\ocaml{|}), sequential (\ocaml{>}$x$\ocaml{>}),
pruning (\ocaml{<}$x$\ocaml{<}), and ``otherwise'' (\ocaml{;})
combinators. The expression $e_1$ \ocaml{|} $e_2$ runs $e_1$ and
$e_2$ in parallel, publishing any value published by either of
them. The expression $e_1$ \ocaml{>}$x$\ocaml{>} $e_2$ binds each
value published by $e_1$ to $x$ and executes $e_2$ for each such
binding, building a concurrent pipeline. The expression $e_1$
\ocaml{<}$x$\ocaml{<} $e_2$ binds the \emph{first} value published
by $e_2$ to $x$ and executes $e_1$. Finally, $e_1$\ocaml{;}$e_2$
runs $e_2$ only if running $e_1$ does not publish any values.

@@

Because of how we defined the syntax of the combinators, and since
parsing with the tool is currently whitespace insensitive, the
multi-character combinators (sequence and pruning) are parsed as
three separate lexical tokens (this is visible in how whitespace
is inserted in the error message above). This means we will also
run into \emph{unresolvable} ambiguities, for example if our base
language includes comparison of numbers with \ocaml{<} and
\ocaml{>}. With such a base language (assuming no associativity
for \ocaml{>}), parsing the expression
%
``\ocaml{42 >x> f(x)}''
%
will result in the following error message:

@@

The change in the first alternative can be done after composition,
and would allow us to reuse both language definitions without
modifications. In the latter two alternatives, we lose reuse of
the definitions of Orc combinators, but place no additional
restrictions on the base language. The preferred resolution
strategy is likely to differ between different compositions and
different languages.
%
For example, another unresolvable ambiguity is going to show up if
the base language uses semi-colons, e.g. for sequencing
expressions (this would clash with the ``otherwise'' combinator).
In this case, there is no reasonable way to disambiguate
``\ocaml{e1; e2}'' without changing the syntax of one of the
operations.

>>>>

\synctex=1
\begin{document}
% TODO: Decide on British or American English.

@@

% TODO: Make sure that forbids are explained

@@

\subsection{Orc} \label{sec:evaluation-orc}
%TODO Title should be descriptive

In this section, we consider the use case of defining a language
and composing it with another previously defined language, showing
how we can deal with the resulting ambiguities that arise from the
composition.
%
The language being defined is a subset of
Orc~\cite{kitchinOrc2009}, a functional programming language which
includes a number of special-purpose combinators for coordinating
concurrent workflows.

In Orc, every expression can ``publish'' zero or more values. Orc
defines four combinators for orchestrating these published values:
the parallel (\ocaml{|}), sequential (\ocaml{>}$x$\ocaml{>}),
pruning (\ocaml{<}$x$\ocaml{<}), and ``otherwise'' (\ocaml{;})
combinators.
%
The expression $e_1$ \ocaml{|} $e_2$ runs $e_1$ and
$e_2$ in parallel, publishing any value published by either of
them.
%
The expression $e_1$ \ocaml{>}$x$\ocaml{>} $e_2$ executes
$[x\mapsto v]e_2$ for \emph{each} value $v$ published by $e_1$,
building a concurrent pipeline.
%
The expression $e_1$ \ocaml{<}$x$\ocaml{<} $e_2$ executes
$[x\mapsto v]e_1$ for \emph{first} value $v$ published by $e_2$
(discarding any remaining values published by $e_2$).
%
Finally, $e_1$\ocaml{;}$e_2$ runs $e_2$ only if running $e_1$ does
not publish any values.

@@

Because of how we defined the syntax of the combinators, and since
parsing with the tool is currently whitespace insensitive, the
multi-character combinators (sequencing and pruning) are parsed as
three separate lexical tokens (this is visible in how whitespace
is inserted in the error message above). This means we can also
run into \emph{unresolvable} ambiguities, for example if our base
language includes comparison of numbers with \ocaml{<} and
\ocaml{>}. With such a base language (assuming no associativity
for \ocaml{>}), parsing the expression
%
``\ocaml{42 >x> f(x)}''
%
will result in the following error message:

@@

\noindent
The change in the first alternative is somewhat ad-hoc, but can be done after composition
and would allow us to reuse both language definitions without
modifications. In the latter two alternatives, we lose reuse of
the definitions of Orc combinators, but place no additional
restrictions on the base language. The preferred resolution
strategy is likely to differ between different compositions and
different languages.
%
For example, another unresolvable ambiguity is going to show up if
the base language uses semi-colons, e.g. for sequencing
expressions (this would clash with the ``otherwise'' combinator).
In this case, there is no reasonable way to disambiguate
``\ocaml{e1; e2}'' without changing the syntax of one of the
operations.

The main takeaway from this case study is that resolvable
ambiguity makes composition of languages less restrictive than if
all ambiguity is completely banned. The approach is strictly more
general since dynamic resolvability analysis allows deferring
disambiguation to the programmer, while removing ambiguities from
the resulting grammar is also possible.
